Philosophy of DS\\ \large Best Practices for the Application of Computational Methods to Questions of Meaning

### 2017-06-01 pds/Paper/paper_pds.tex
made progress. certainly. coke is the best.

%\usepackage{biblatex}
%\addbibresource{pds_references.bib}
\usepackage{csquotes}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of DS (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbours the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants to point to one particular attempt to provide general-purpose guidelines for ML and draw on these existing insights to suggest guidelines and best practices for computational methods in DS.
As the domain of DS differs in important aspects from other applications of ML, notably also for Natural Language Processing (NLP),
this involves assessing  the applicability of general purpose ML guidelines to DS with these difference in mind.
\end{abstract}
\section{Introduction}
%1-2 paragraphs, talk about application, overall field
\paragraph{Place of Machine Learning in society}
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications. 
\cite{algorithms2016}
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

\paragraph{Harmful Effects}
There are several different sources of harmful effects that these techniques can bring about. With the very large data collected, stored and processed, issues of privacy arise naturally. Especially when data mining techniques are used to scrape the web, people's explicit consent to provide their data can easily go by the board. Another recurrent problem is the two-sided nature of any technology: it can be used for the good or for dubious purposes, as fraud or spying, for example. Finally, big data and ML approaches brings the risk of reenforcing structural bias along racial, gender or others social lines in society. Models trained on biased historical data, for example, have been used to determine prison sentences which lead to consistently biased outcomes against blacks and latinos in the United States of America, land of the brave and home of the free \cite{angwin2016machine}.

In the present essay, we will mostly be concerned with the last of these problematic issues surrounding ML.

%\begin{quote}
%In the years since, several disparate impact cases have made their way to the Supreme Court and lower courts, most having to do with employment discrimination. This June, the Supreme Court’s decision in Texas Dept. of Housing and Community Affairs v. Inclusive Communities Project, Inc. affirmed the use of the disparate impact theory to fight housing discrimination. The Inclusive Communities Project had used a statistical analysis of housing patterns to show that a tax credit program effectively segregated Texans by race. Sorelle Friedler, a computer science 
%researcher at Haverford College and a fellow at Data \& Society, called the Court’s decision “huge,” both “in favor of civil rights…and in favor of statistics.”
%\end{quote}

%\paragraph{Contribution}
%largest part of intro, preview of paper, methodology!
\paragraph{Public Attention is in ML minor extent in NLP, not in DS}
Many of the issues lately mentioned are of great importance to modern society at large, especially as. Consequently, people working at the intersection of information technology and ethics have attempted to lay down foundations on how to handle these technologies. Many of these efforts have naturally been directed at the broad topic of ML in general.\cite{leidner2017ehtical} Only a few years ago the discussion has turned on the specific use of ML for analysis and modeling of language - often referred to as \emph{natural language processing}. Presumably because this is only one of many particular fields where ML is applied and the discussion of ethics  is bound to pertain to less cases of applications throughout society, it took some time for the discussion to reach this field.

A recently as 2016, however, a discussion has opened up in a hitherto completely unmentioned domain. ML techniques used for the computer-based study of language meaning - a field known as \emph{computational DS} - has come under scrutiny.
Various studies have shown that these methods yield results which are in important ways biased against particular groups as, for example, women \cite{google} \cite{wagner2015s}.
\textbf{Why this is important. Mention Google searches using semantic information.}

The realization that algorithms fed on big data come with a host of problematic issues has led to the emergence of a variety of principles, concrete guidelines and public as well as institutional scrutiny.

\paragraph{Contribution: Drawing on ML discussion for DS}
In the present paper we want to draw on the existing efforts made by experts in the fields of ethics, computer science, public policy and law to extend these principles and guidelines to the field of DS. Specifically, we will use the \emph{Principles for Accountable Algorithms} (PAA) issued by the organization \emph{Fairness, Accountability and Transparency in Machine Learning} (FATML) - a consortium of researches and experts from private and public organizations working on ML. 
The PAA comprise five guiding principles: responsibility, explainability, accuracy, auditibility and fairness \cite{principles}.
We will assess potential violation of techniques used in DS for each of these principles and, where applicable, issue guidelines to prevent their negative impact upon vulnerable groups.

Currently, we want to focus on aspects of discrimination and bias in the outcome of DS models for two reasons. For one, these are the least explored ethical issues for DS. Many of the ideas, guidelines and principles concerning privacy, double use or economic impact are directly transferable from the general ML debate to any sub-field that similarly employs big data and algorithms. Computational DS is no exception. Debates concerning privacy etc. can follow the same lines as the discussion to establish general purpose ethical principles for ML. The issue of discrimination and bias, however, has a special relation to DS because the field occupies a unique position of computational language research. The central concern of DS, as its name gives away, is the concept of meaning. As we will argue presently, this relates it in a special way to many sensitive topics in the general ML discussion. Put succinctly, it sits at the intersection of concerns pertaining to data selection, validity or questionability of computational methods, interpretation of results and usage thereof.

\paragraph{Methodoology}
The method employed by the present contribution is one of scanning previous efforts in a similar vein, identifying useful parts and applying them to ethical questions surrounding DS. There will be no foundational efforts undertaken to sustain the validity of the principles thus transferred.  Their validity is presupposed. The analysis contained in this contribution merely serves to discuss the usefulness of the principles in particular domain, and assert it where due.

\paragraph{Paper Overview}
The paper is structured as follows: In \hyperlink{sec2}{Section 2} we first give a very brief description of ML,  and DS as well as their relation to each other. Especially for DS we want to provide overview over societal domains where such technology is currently applied and could conceivably be applied in the future. \hyperlink{sec3}{Section 3} contains a discussion of the important ways in which DS, as a particular application of big data analysis, differs from other instances of ML. For the present paper these particularities are important insofar as they entail differences when it comes to its risks and accountabilities across the spectrum of ML. In \hyperlink{sec4}{Section 4}, then, we try to apply the \emph{Principles for Accountable Algorithms and a Social Impact Statement for Algorithms} by FAT/ML to the specific case of DS. \hyperlink{sec5}{Section 5} contains concluding remarks while \hyperlink{sec6}{Section 6} lists the references.
\section{Machine Learning, Natural Language Processing and Distributional Semantics}\hypertarget{sec2}{ }
\paragraph{What is ML?}
The term \emph{Machine Learning} is an umbrella term for a variety of statistical methods that are able to \emph{learn} from a series of examples presented to them. The the set of examples - often called the \emph{training data} - is not simply memorized but the ML algorithm extracts generalizable knowledge \cite{domingos2012few}.
 What the algorithm learns are patterns and relationships within the training data and the result of the learning process is a statistical description - or \emph{model} - of the training data \cite{fayyad2001digital}.
\paragraph{What is NLP?}
Natural language processing is a field of research that uses computational tools to build natural language text or speech applications. It is related to computational linguistic, an area of linguistic that uses computational techniques to get a scientific understanding of natural language. The knowledge thus obtained can be used in  to develop computer systems that serve a wide variety of purposes. Principal applications include translation between natural languages, artificial intelligence, user interfaces, like chat-bots, for example, summarization of text or speech recognition \cite{chowdhury2003natural}.
\paragraph{Relation: ML - NLP}
The area of  frequently serves itself of methods from ML, for example to process very large textual corpora. Indeed, parts of  can be thought of as particular applications of ML techniques to the domain of natural language.
\paragraph{What is DS?}
DS is a linguistic theory of meaning looking to represent lexical meaning of expressions as a function of the context in which they appear. Its theoretically foundation is the so-called \emph{distributional hypothesis} which holds that two expressions are similar in meaning, if they occur in similar contexts \cite{harris1954distributional}.
%Althought it was originally developed as a distributional theory - in the sense that ... - in the middle of the 20th century, it can be seen as following the footsteps of structural linguists, notably Bloomfield \cite{bloomfield}

@@

@@

The lesson of this particular successfull debiasing endeavour is straightforward. Performance is no general excuse for designers of systems employing DS to work with meanings that reflect societal bias. Efforts can be made to process the DS model after its original training to rid it of discriminatory disparaties. In general domain ML there is already a sizeable body of literature of finding an removing so-called \emph{disparate impact} - unintentional discrimination (cf. \cite{feldman2015certifying} \cite{dwork2012fairness}).

Similar collections of standardized techniques should be developed to debias semantic spaces. Doing this is especially important, of course, if the DS model to be used for some outside application. 
Perhaps a complement to the attempts to debiase embeddings, then, are the studies whose stated goal it is to uncover biases in corpora, as eg. \cite{wagner2015s} or \cite{herbelot2012distributional}. Findings that indicate clear instances of bias in widely used corpora can then be used as an indication where  debiasing would have to take place before the meanings should be put to further use.

Both of these

>>>>

%\usepackage{biblatex}
%\addbibresource{pds_references.bib}
\usepackage{csquotes}
\begin{document}
\maketitle
\begin{abstract}
The increase in availability of data and advances in statistical methods has led to the use of Machine Learning (ML) technology in an immense variety of societal domains. ML algorithms are used to inform decisions as diverse and life-changing as access to financial credit, incarceration, university admission, medical treatment or the legal boundaries of espionage by the NSA.
The use of this technology is not always unproblematic, as its careless application can cause serious harm, even if such effects occur unintentionally.
The seriousness of the potential to disrespect people's privacy and integrity or systematically disadvantage specific groups has led to a recent debate concerning ethical guidelines and best practices for the general use of ML.
Little has been said, however, about the proper use of algorithms in the specific field of Distributional Semantics (DS). Some researchers have called attention to possible dangers of using ML in DS, as it harbours the potential to reinforce harmful biases along dimensions of race and gender.
The present paper wants to point to one particular attempt to provide general-purpose guidelines for ML and draw on these existing insights to suggest guidelines and best practices for computational methods in DS.
As the domain of DS differs in important aspects from other applications of ML, notably also for Natural Language Processing (NLP),
this involves assessing  the applicability of general purpose ML guidelines to DS with these difference in mind.
\end{abstract}
\section{Introduction}
%1-2 paragraphs, talk about application, overall field
As vast amounts of digital data have become available over past decades, algorithms are employed to use the data in order to take and inform decisions all across society. 
Algorithms have a say in determining who is granted mortgages, who is released on bail or invited for job applications \cite{algorithms2016}.
The theory behind many of these algorithms is \emph{machine learning} (ML).
Although most inventors and designers of ML algorithms certainly do not harbor malicious intent, the risk of unwillingly harming people is real.

There are several different sources of harmful effects that these techniques can bring about. With the very large data collected, stored and processed, issues of privacy arise naturally. Especially when data mining techniques are used to scrape the web, people's explicit consent to provide their data can easily go by the board. Another recurrent problem is the two-sided nature of any technology: it can be used for the good or for dubious purposes, as fraud or spying, for example. Finally, big data and ML approaches brings the risk of reenforcing structural bias along racial, gender or others social lines in society. Models trained on biased historical data, for example, have been used to determine prison sentences which lead to consistently biased outcomes against blacks and latinos in the United States of America \cite{angwin2016machine}.

%largest part of intro, preview of paper, methodology!
Clearly, many of these issues are of great importance to modern society at large. Consequently, people working at the intersection of information technology and ethics have attempted to lay down foundations on how to handle these technologies. The realization that algorithms fed on big data come with a host of problematic issues has led to the emergence of a variety of principles, concrete guidelines and public as well as institutional scrutiny. Many of these efforts have naturally been directed at the broad topic of ML in general.\cite{leidner2017ehtical} Only a few years ago the discussion has turned on the specific use of ML for analysis and modeling of language - sometimes broadly referred to as \emph{natural language processing} (NLP). This is only one of many particular fields where ML is applied and the discussion of ethics in ML naturally dissipated from more general to more domain-specific applications.

A recently as 2016, however, a discussion has opened up in a hitherto completely undiscussed domain. ML techniques used for the computer-based study of language meaning - a field known as \emph{computational DS} - has come under scrutiny.
Various studies have shown that these methods can  yield results which are in important ways biased against particular groups as, for example, women \cite{google} \cite{wagner2015s}.
\textbf{Why this is important. Mention Google searches using semantic information.}

In the present paper we want to draw on the existing efforts made by experts in the fields of ethics, computer science, public policy and law to extend principles  and guidelines for general purpose ML ethics to the field of DS. Specifically, we will use the \emph{Principles for Accountable Algorithms} (PAA) issued by the organization \emph{Fairness , Accountability and Transparency in Machine Learning} (FATML) - a consortium of researches and experts from private and public organizations working on ML. 
The PAA comprise five guiding principles: responsibility, explainability, accuracy, auditibility and fairness \cite{principles}.
We will assess potential violation of techniques used in DS for each of these principles and, where applicable, issue guidelines to prevent their negative impact upon vulnerable groups. 
The present contribution works by scanning previous efforts made in a similar vein, identifying useful parts therein and applying them to ethical questions surrounding DS. We will make no foundational efforts to sustain the validity of the principles. They are presupposed. Our analysis merely serves to discuss the usefulness of the principles in particular domain, and assert it where due.

Currently, we want to focus on aspects of discrimination and bias in the outcome of DS models. This for two reasons: For one, these are the least explored ethical issues for DS. Many of the ideas, guidelines and principles concerning privacy, double use or economic impact are directly transferable from the general ML debate to any sub-field that similarly employs big data and algorithms. Computational DS is no exception. Debates concerning privacy etc. can follow the same lines as the discussion to establish general purpose ethical principles for ML. The issue of discrimination and bias, however, has a special relation to DS because the field occupies a unique position of computational language research. The central concern of DS, as its name gives away, is the concept of meaning. As we will argue presently, this relates it in a special way to many sensitive topics in the general ML discussion. Put succinctly, it sits at the intersection of concerns pertaining to data selection, validity or questionability of computational methods, interpretation of results and usage thereof.

The paper is structured as follows: In \hyperlink{sec2}{Section 2} we first give a very brief description of ML,  and DS as well as their relation to each other. Especially for DS we want to provide overview over societal domains where such technology is currently applied and could conceivably be applied in the future. \hyperlink{sec3}{Section 3} contains a discussion of the important ways in which DS, as a particular application of big data analysis, differs from other instances of ML. For the present paper these particularities are important insofar as they entail differences when it comes to its risks and accountabilities across the spectrum of ML. In \hyperlink{sec4}{Section 4}, then, we try to apply the \emph{Principles for Accountable Algorithms} by FATML to the specific case of DS. \hyperlink{sec5}{Section 5} contains concluding remarks while \hyperlink{sec6}{Section 6} lists the references.
\section{Machine Learning and Distributional Semantics}\hypertarget{sec2}{ }
The term \emph{machine learning} is an umbrella term for a variety of statistical methods that are able to \emph{learn} from a series of examples presented to them. The set of examples - often called the \emph{training data} - is not simply memorized, but the ML algorithm extracts generalizable knowledge like structures, patterns or relationships \cite{domingos2012few}.
The result of the learning process is a statistical description - or \emph{model} - of the training data \cite{fayyad2001digital}.

Natural language processing (NLP) is a field of research that uses computational tools to build natural language text or speech applications. It is related to computational linguistic, an area of linguistic that uses computational techniques to get a scientific understanding of natural language. The knowledge thus obtained can be used in  to develop computer systems that serve a wide variety of purposes. Principal applications include translation between natural languages, artificial intelligence, user interfaces, like chat-bots, for example, summarization of text or speech recognition \cite{chowdhury2003natural}.
\paragraph{Relation: ML - NLP}
The area of  frequently serves itself of methods from ML, for example to process very large textual corpora. Indeed, parts of  can be thought of as particular applications of ML techniques to the domain of natural language.
\paragraph{What is DS?}
DS is a linguistic theory of meaning looking to represent lexical meaning of expressions as a function of the context in which they appear. Its theoretically foundation is the so-called \emph{distributional hypothesis} which holds that two expressions are similar in meaning, if they occur in similar contexts \cite{harris1954distributional}.
%Althought it was originally developed as a distributional theory - in the sense that ... - in the middle of the 20th century, it can be seen as following the footsteps of structural linguists, notably Bloomfield \cite{bloomfield}

@@

A further point to consider is the fact that the nature of the object supposedly captured by DS - meaning - is highly contested. Conceptual confusion and debate abound concerning the exact relation around the concepts of meaning, groundedness, speaker and community. This can make it even harder to adequately respond to calls of responsibility in these matters because the relation that holds between corpus, DS techniques, meaning and society are far from settled.

@@

The lesson of this particular successfull debiasing endeavour is straightforward. Performance is no general excuse for designers of systems employing DS to work with meanings that reflect societal bias. Efforts can be made to process the DS model after its original training to rid it of discriminatory disparaties. In general domain ML there is already a sizeable body of literature of finding an removing so-called \emph{disparate impact} - unintentional discrimination (cf. \cite{feldman2015certifying} \cite{dwork2012fairness}). 
Similar collections of standardized techniques should be developed to debias semantic spaces. Doing this is especially important, of course, if the DS model to be used for some outside application. 
Perhaps a complement to the attempts to debiase embeddings, then, are the studies whose stated goal it is to uncover biases in corpora, as eg. \cite{wagner2015s} or \cite{herbelot2012distributional}. Findings that indicate clear instances of bias in widely used corpora can then be used as an indication where  debiasing should have to take place before the DS is put to further use.
Thus, it appears that a two-fold effort is required in to prevent damage from happening in the employment of algorithmically generated meanings. On the one hand, one will constantly have to scan the semantic spaces gathered from corpora for undesired connotations or other semantic relationships within it; on the other, once such biases are certified, efforts need to be undertaken, so that systems using this semantic information do not systematically disadvantage people.

In this fashion fairness principle could be addressed in DS. This does not yet respond, however, to the question where exactly responsibility lies - that is, with whom - in case something does go wrong.
